---
title: "Environmental attitudes in Europe"
description: "European Values Study Wave 5 (2017-2018) and the 'likert' package"
author: "Marta Kołczyńska"
date: 2019-08-28T12:06:37
categories: ["R"]
tags: ["surveys", "EVS", "R", "FridaysForFuture", "environmental attitudes", "environment", "likert"]
output: 
  blogdown::html_page:
    toc: true
    toc_depth: 2
---

The climate protests in March 2019 mobilized over a million of people around the globe. A team of social scientists from universities across Europe organized a survey of the #FridaysForFuture strike events on March 15 in 13 cities in nine countries. The report can be found [here](https://osf.io/xcnzh/){target="_blank"}. A new wave of climate protests (and surveys) is planned for the end of September. 

Naturally, most participants at these protests are acutely aware of the environmental threats and motivated to take action. But what is the situation like among the general population?

Here is a quick summary with data from the [European Values Study](https://europeanvaluesstudy.eu/){target="_blank"} wave 5 (from the 2nd pre-release in July 2019), using the [`likert` package](https://github.com/jbryer/likert){target="_blank"}.

```{r setup, warning=FALSE, message=FALSE, echo = FALSE}
library(plyr) # for manipularing data (mapvalues)
library(tidyverse) # for data wrangling
library(countrycode) # for converting between country codes and names
library(likert) # for plotting likert-type items
```

```{r data-prep, warning=FALSE, message=FALSE, echo = FALSE, eval = FALSE}
evs_5_env <- import("C:/Users/mkolc/Google Drive/Work in progress/R playground/data-for-harm/EVS/ZA7500_v2-0-0.sav.zip", 
                user_na = TRUE) %>%
  filter(c_abrv %in% c("AT", "CH", "DE", "IT", "NL", "PL", "SE", "GB"), gweight > 0)
rio::export(evs_5_env, "environment_evs5.csv")
```


```{r likert-plot-cntry, warning=FALSE, message=FALSE, echo = FALSE, fig.width = 9, fig.height = 12}
evs_5_env <- read.csv("data/environment_evs5.csv", stringsAsFactors = TRUE)  %>%
  select(cntry = c_abrv, v199,v200,v201,v202,v203) %>% 
  mutate_at(vars(v199,v200,v201,v202,v203), funs(ifelse(. < 0, NA, .))) %>%
  mutate_at(vars(v199,v200,v201,v202,v203), funs(plyr::mapvalues(., c(1,2,3,4,5), c(5,4,3,2,1)))) %>%
  mutate(country = countrycode(cntry, "iso2c", "country.name")) %>% 
  mutate_at(vars(v199,v200,v201,v202,v203), funs(ifelse(. < 0, NA, .))) %>%
  mutate_at(vars(v199,v200,v201,v202,v203), funs(ordered(., levels = c(1,2,3,4,5),
                                                         labels = c("disagree strongly", 
                                                                    "disagree", 
                                                                    "neither agree nor disagree", 
                                                                    "agree", 
                                                                    "agree strongly")))) %>%
  rename('I would give part of my income if I were certain that the money would be used to prevent environmental pollution' = v199,
         'It is just too difficult for someone like me to do much about the environment' = v200,
         'There are more important things to do in life than protect the environment' = v201,
         'There is no point in doing what I can for the environment unless others do the same' = v202,
         'Many of the claims about environmental threats are exaggerated' = v203)
  

data_env <- evs_5_env %>% select(-cntry, -country)
cntry <- select(evs_5_env, country)

data_env_lg <- likert::likert(data_env, grouping = cntry$country)
plot(data_env_lg) + theme_bw(12)
```
